### 2024-04-18 docs/SOSP 2024/motivation.tex
Edit figures and references

\subsection{Motivating Example}
\begin{figure}[htb]
  \centering
  \includegraphics[width=\linewidth]{figures/HIR.PNG}
  \caption{The representation of a model in high-level IR and the model 
  after trees are padded and reordered.\TODO{drop reorder, make 1,2 depth 1 and 3,4 depth 2.} }
  \label{Fig:HIRExample}
\end{figure}

Consider a model with four trees, two 
trees of depth 1 and two of depth 2 (Figure \ref{Fig:HIRExample}).
% We use this model as our running example throughout this section.
We first describe a simple schedule that processes one tree at a time 
for all input rows. The scheduling constructs are fairly intuitive. We 
defer a detailed explanation of the scheduling language to Section~\ref{sec:scheduling}. 
The loop over the trees is split into two -- one that
iterates over the first two trees (Trees 1 and 2 with depth 1) and 
the second that iterates over the last two trees (Trees 3 and 4 with
depth 2). Straightforward traversal of trees requires a \op{while} loop
and involves branching to check if a leaf has been reached. 
One way to avoid this as done in this schedule is to unroll the 
tree walks for each tree. 
When the schedule specifies that the tree walks should be unrolled, 
\Treebeard{} pads the trees so that all leaves are at the same depth (the padded trees are also shown in Figure~\ref{Fig:HIRExample}).

\begin{lstlisting}[style=c++]
  reorder(tree, batch)
  // Fiss tree loop so trees with equal depth 
  // are processed together  
  split(tree, t_depth1, t_depth2, 2)
  // Unroll the tree walks
  unrollWalk(t_depth1, 1)
  unrollWalk(t_depth2, 2)
\end{lstlisting}
The concrete implementation of this schedule (in \Treebeard{}'s IRs) 
is as follows.
\begin{lstlisting}[style=c++]
  model = ensemble(...)
  for t_depth1 = 0 to 2 step 1 {
    T = getTree(ensemble, t_depth1)
    for batch = 0 to BATCH_SIZE step 1 {
      treePred = walkDecisionTree(T, 
                    input[batch]) <unrollDepth = 1>
      reduce(result[batch], treePred)
    }
  }
  for t_depth2 = 2 to 4 step 1 {
    T = getTree(ensemble, t_depth2)
    for batch = 0 to BATCH_SIZE step 1 {
      treePred = walkDecisionTree(T,
                    input[batch]) <unrollDepth = 2>
      reduce(result[batch], treePred) <'+', 0.0>
    }
  }
\end{lstlisting}
This schedule is ideally suited for a single-core CPU. It maximizes 
the reuse of trees in the L1 cache. However, it doesn't exploit  
any parallelism. 
%% RG, 
%%  and is therefore ill-suited for parallel processors.
% Further, the \op{batch} for loop can be tiled to get locality benefits.

One form of parallelism that can be exploited is to process rows in 
parallel. 
%% RG:
While this may work for multi-core CPUs\footnote{As we report in Section~\ref{sec:results}, other ways to parallelize can benefit inference on CPUs as well.}, with massively parallel processors like GPUs,
 this strategy may not yield sufficient parallel work. To expose more parallelism, 
we can additionally parallelize across trees as done in the schedule below. 
%% RG:
% With GPUs, the performance of generated code would also vary with the 
% kernel configuration, which specifies the number of thread blocks\footnote{We 
% use NVIDIA CUDA terminology in the discussion, but as we demonstrate in 
% Section~\ref{sec:results}, our framework can generate code for AMD GPUs as well.}, 
% the size of thread blocks and so on. 
% A possible strategy to accomplish this is encoded in the following schedule.
\begin{lstlisting}[style=c++]
  // Split the trees into two sets
  tile(tree, t0, t1, 2)
  reorder(batch, t1, t0)
  // Fiss loop so that trees with equal 
  // depth are processed together
  split(t0, t0_depth1, t0_depth2, 2)
  unrollWalk(t0_depth1, 1)
  unrollWalk(t0_depth2, 2)
  // Configure the GPU kernel dimensions
  gpuDimension(batch, grid.x)
  gpuDimension(t1, block.x)
\end{lstlisting}
This schedule generates an inference function that runs on the GPU. 
The inference routine processes one input row per thread block (since the \op{batch}
loop is mapped directly to \op{grid.x}).
It also splits the trees into two sets by tiling the \op{tree} loop.
Each of the two sets is processed in parallel. We unroll the tree walks 
for each tree. The IR generated is as follows. 
\begin{lstlisting}[style=c++]
  model = ensemble(...)
  par.for batch = 0 to BATCH_SIZE step 1 <grid.x> {
    par.for t1 = 0 to 2 step 1 <block.x> {
      for t0_depth1 = 0 to 2 step 2 {
        T = getTree(ensemble, t0_depth1 + t1)
        treePred = walkDecisionTree(T, 
                        input[batch]) <unrollDepth = 1>
        reduce(result[batch], treePred)
      }
      for t0_depth2 = 2 to 4 step 2 {
        T = getTree(ensemble, t0_depth2 + t1)
        treePred = walkDecisionTree(T,
                        input[batch]) <unrollDepth = 2>
        reduce(result[batch], treePred) <'+', 0.0>
      }
    }
  }
\end{lstlisting}

% In the case of this schedule, the \op{reduce}
% operation needs special consideration.  
% %% RG
% %% to correctly generate code for this schedule, the compiler needs 
% To generate the correct code for this schedule, the compiler needs to determine
% the required synchronization and temporary storage for reductions 
% given the parallelization strategy specified. In the given example, it 
% has to identify that parallel iterations of the \op{t1} 
% loop accumulate into the same element of the \op{result} array.
% One possible solution is to rewrite the reduction so that each parallel 
% iteration accumulates into a different array element by introducing 
% a temporary buffer (\op{temp}) as follows.
% \begin{lstlisting}[style=c++]
%   float temp[2][BATCH_SIZE]
%   model = ensemble(...)
%   par.for batch = 0 to BATCH_SIZE step 1 <grid.x> {
%     par.for t1 = 0 to 2 step 1 <block.x> {
%       for t0_depth1 = 0 to 2 step 2 {
%         T = getTree(ensemble, t0_depth1 + t1)
%         treePred = walkDecisionTree(T, 
%                       input[batch]) <unrollDepth = 1>
%         reduce(temp[t1][batch], treePred)
%       }
%       for t0_depth2 = 2 to 4 step 2 {
%         T = getTree(ensemble, t0_depth2 + t1)
%         treePred = walkDecisionTree(T,
%                       input[batch]) <unrollDepth = 2>
%         reduce(temp[t1][batch], treePred) <'+', 0.0>
%       }
%     }
%     result[batch] = reduce_dimension(temp[:][batch], 0)
%   }
% \end{lstlisting}
% Here, partial results are accumulated into \op{temp} and then
% reduced across the \op{t1} dimension to get the final result.
Note that while unrolling helps avoid branching, it likely increases 
the total amount of computation. Another option is to not unroll 
and let the GPU manage the branching. The schedules with and 
without unrolling place different constraints on the target processor,
and the best choice depends on the characteristics of the model
and micro-architectural features like register file size, handling 
of branch divergence etc.~\cite{FungMicro11, SomethingElse}.

>>>>

\subsection{Motivating Example}
\begin{figure}[htb]
  \centering
  \includegraphics[width=\linewidth]{figures/HIR.PNG}
  \caption{Model for motivating example. The model has four trees, two of depth 1 and two of depth 2.}
  \label{Fig:HIRExample}
\end{figure}

Consider a model with four trees, two 
trees of depth 1 and two of depth 2 (\circled{A} in Figure \ref{Fig:HIRExample}).
% We use this model as our running example throughout this section.
We first describe a simple schedule that processes one tree at a time 
for all input rows. The scheduling constructs are fairly intuitive. We 
defer a detailed explanation of the scheduling language to Section~\ref{sec:schedule}. 
The loop over the trees is split into two -- one that
iterates over the first two trees (Trees 1 and 2 with depth 1) and 
the second that iterates over the last two trees (Trees 3 and 4 with
depth 2). Straightforward traversal of trees requires a \op{while} loop
and involves branching to check if a leaf has been reached. 
One way to avoid this as done in this schedule is to unroll the 
tree walks for each tree. 
When the schedule specifies that the tree walks should be unrolled, 
\Treebeard{} pads the trees so that all leaves are at the same depth
(\circled{B} in Figure~\ref{Fig:HIRExample}).

\begin{lstlisting}[style=c++]
  reorder(tree, batch)
  // Fiss tree loop so trees with equal depth 
  // are processed together  
  split(tree, t_depth1, t_depth2, 2)
  // Unroll the tree walks
  unrollWalk(t_depth1, 1)
  unrollWalk(t_depth2, 2)
\end{lstlisting}
The concrete implementation of this schedule (in \Treebeard{}'s IRs) 
is as follows.
\begin{lstlisting}[style=c++]
  model = ensemble(...)
  for t_depth1 = 0 to 2 step 1 {
    T = getTree(model, t_depth1)
    for batch = 0 to BATCH_SIZE step 1 {
      treePred = walkDecisionTree(T, 
                    input[batch]) <unrollDepth = 1>
      reduce(result[batch], treePred)
    }
  }
  for t_depth2 = 2 to 4 step 1 {
    T = getTree(model, t_depth2)
    for batch = 0 to BATCH_SIZE step 1 {
      treePred = walkDecisionTree(T,
                    input[batch]) <unrollDepth = 2>
      reduce(result[batch], treePred) <'+', 0.0>
    }
  }
\end{lstlisting}
This schedule is ideally suited for a single-core CPU. It maximizes 
the reuse of trees in the L1 cache. However, it doesn't exploit  
any parallelism. 
%% RG, 
%%  and is therefore ill-suited for parallel processors.
% Further, the \op{batch} for loop can be tiled to get locality benefits.

One form of parallelism that can be exploited is to process rows in 
parallel. 
%% RG:
While this may work for multi-core CPUs\footnote{As we report in Section~\ref{sec:results}, other ways to parallelize can benefit inference on CPUs as well.}, with massively parallel processors like GPUs,
 this strategy may not yield sufficient parallel work. To expose more parallelism, 
we can additionally parallelize across trees as done in the schedule below. 
%% RG:
% With GPUs, the performance of generated code would also vary with the 
% kernel configuration, which specifies the number of thread blocks\footnote{We 
% use NVIDIA CUDA terminology in the discussion, but as we demonstrate in 
% Section~\ref{sec:results}, our framework can generate code for AMD GPUs as well.}, 
% the size of thread blocks and so on. 
% A possible strategy to accomplish this is encoded in the following schedule.
\begin{lstlisting}[style=c++]
  // Split the trees into two sets
  tile(tree, t0, t1, 2)
  reorder(batch, t1, t0)
  // Fiss loop so that trees with equal 
  // depth are processed together
  split(t0, t0_depth1, t0_depth2, 2)
  unrollWalk(t0_depth1, 1)
  unrollWalk(t0_depth2, 2)
  // Configure the GPU kernel dimensions
  gpuDimension(batch, grid.x)
  gpuDimension(t1, block.x)
\end{lstlisting}
This schedule generates an inference function that runs on the GPU. 
The inference routine processes one input row per thread block (since the \op{batch}
loop is mapped directly to \op{grid.x}).
It also splits the trees into two sets by tiling the \op{tree} loop.
Each of the two sets is processed in parallel. We unroll the tree walks 
for each tree. The IR generated is as follows. 
\begin{lstlisting}[style=c++]
  model = ensemble(...)
  par.for batch = 0 to BATCH_SIZE step 1 <grid.x> {
    par.for t1 = 0 to 2 step 1 <block.x> {
      for t0_depth1 = 0 to 2 step 2 {
        T = getTree(model, t0_depth1 + t1)
        treePred = walkDecisionTree(T, 
                        input[batch]) <unrollDepth = 1>
        reduce(result[batch], treePred)
      }
      for t0_depth2 = 2 to 4 step 2 {
        T = getTree(model, t0_depth2 + t1)
        treePred = walkDecisionTree(T,
                        input[batch]) <unrollDepth = 2>
        reduce(result[batch], treePred) <'+', 0.0>
      }
    }
  }
\end{lstlisting}

% In the case of this schedule, the \op{reduce}
% operation needs special consideration.  
% %% RG
% %% to correctly generate code for this schedule, the compiler needs 
% To generate the correct code for this schedule, the compiler needs to determine
% the required synchronization and temporary storage for reductions 
% given the parallelization strategy specified. In the given example, it 
% has to identify that parallel iterations of the \op{t1} 
% loop accumulate into the same element of the \op{result} array.
% One possible solution is to rewrite the reduction so that each parallel 
% iteration accumulates into a different array element by introducing 
% a temporary buffer (\op{temp}) as follows.
% \begin{lstlisting}[style=c++]
%   float temp[2][BATCH_SIZE]
%   model = ensemble(...)
%   par.for batch = 0 to BATCH_SIZE step 1 <grid.x> {
%     par.for t1 = 0 to 2 step 1 <block.x> {
%       for t0_depth1 = 0 to 2 step 2 {
%         T = getTree(ensemble, t0_depth1 + t1)
%         treePred = walkDecisionTree(T, 
%                       input[batch]) <unrollDepth = 1>
%         reduce(temp[t1][batch], treePred)
%       }
%       for t0_depth2 = 2 to 4 step 2 {
%         T = getTree(ensemble, t0_depth2 + t1)
%         treePred = walkDecisionTree(T,
%                       input[batch]) <unrollDepth = 2>
%         reduce(temp[t1][batch], treePred) <'+', 0.0>
%       }
%     }
%     result[batch] = reduce_dimension(temp[:][batch], 0)
%   }
% \end{lstlisting}
% Here, partial results are accumulated into \op{temp} and then
% reduced across the \op{t1} dimension to get the final result.
Note that while unrolling helps avoid branching, it likely increases 
the total amount of computation. Another option is to not unroll 
and let the GPU manage the branching. The schedules with and 
without unrolling place different constraints on the target processor,
and the best choice depends on the characteristics of the model
and micro-architectural features like register file size, handling 
of branch divergence etc.~\cite{FungMicro, MilindDivergence}.
